---
title: "Assessment"
author: "Eleanor Harrison"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Assessing the relationship between living standards and chronic asthma

# Introduction

Asthma is a common condition worldwide that can be caused by prolonged indoor exposure to mould (World Health Organization, 2024). Currently, there is a lack of accurate data on the prevalence of mould in social housing, potentially making this association more serious than expected (The Scottish Parliament, 2023).

This exploratory data analysis will investigate whether living standards could have an impact on the prevalence of chronic asthma. To do this, living standards will be defined by SIMD due to the association of social housing with mould, and prevalence of chronic asthma will be measured by the number of prescriptions of drugs used in common preventer inhalers as these are often prescribed when long or short acting bronchodilators are ineffective, suggesting severe chronic asthma.

### Data chosen for Analysis

**SIMD**

SIMD Data is released every 4 years, the dataset used will be the most recent data from 2020 (Public Health Scotland: <https://www.opendata.nhs.scot/dataset/scottish-index-of-multiple-deprivation/resource/acade396-8430-4b34-895a-b3e757fa346e>)

**Prescriptions**

The prescriptions assessed are those for salbutamol, beclometasone, ciclesonide, fluticasone and mometasone as they are the most commonly prescribed drugs used in preventative inhalers (Vincent, 2024)

Prescriptions data used will be *all* months from 2020 to correspond with SIMD year in order to accurately compare the two (Public Health Scotland: <https://www.opendata.nhs.scot/dataset/prescriptions-in-the-community>)

**GP Practices**

Data on GP Practices was also taken from 2020 to ensure datazone areas were up to date (Public Health Scotland: <https://www.opendata.nhs.scot/dataset/gp-practice-contact-details-and-list-sizes/resource/42391720-7dcb-48a2-8070-b9d63b246ac6>)

GP Practice Shapefile Data was extracted from Spatial Hub with the locations updated as of 2020  (<https://data.spatialhub.scot/dataset/gp_practices-is/resource/8389fd1d-563d-4c05-9833-26d9f07fd6cd>)

GP Practice Healthboard Shapefile was extracted from spatialdata.gov.scot
(<https://spatialdata.gov.scot/geonetwork/srv/eng/catalog.search#/metadata/f12c3826-4b4b-40e6-bf4f-77b9ed01dc14>)


```{r libraries, include=FALSE}
library(tidyverse)
library(janitor)
library(knitr)
library(here)
library(gt)
library(sf)
```

### Loading Datasets

```{r loading data, message = FALSE}
#loading datasets with desired variables selected
#some variables have been renamed for continuity between datasets
#clean names also used for continuity
simd2020 <- read_csv(here("data/simd2020.csv")) %>% 
  clean_names() %>% 
  select(data_zone, simd2020v2country_decile) %>% 
  mutate(simd2020v2country_decile = as.factor(simd2020v2country_decile))

gp_practices <- read_csv(here("data/practice_contactdetails_oct2020-open-data.csv")) %>% 
  rename(practice_name = GPPracticeName) %>% 
  clean_names() %>% 
  select(practice_code, data_zone, hb, practice_list_size, practice_name, gp_cluster, postcode)

#prescriptions data taken from jan-dec 2020, 
files <- list.files(here("data", "data_2020"), pattern = "csv")
all_prescriptions_2020 <- files %>% 
  map_dfr(~read_csv(here("data", "data_2020", .)))

# filtering prescriptions for desired drugs and selecting desired columns
inhaler_prescriptions <- all_prescriptions_2020 %>% 
  rename(HB = HBT) %>%
  rename(practice_code = GPPractice) %>% 
  clean_names() %>% 
  select(practice_code, bnf_item_description, number_of_paid_items)%>% 
  filter(str_detect(bnf_item_description, "SALBUTAMOL|BECLOMETASONE|CICLESONIDE|FLUTICASONE|MOMETASONE"))
```

# Comparison of Prescriptions per GP Practice by SIMD

The three datasets used for this analysis (GP practice, SIMD and Prescriptions data) differ by the variables used for location markers therefore cannot be joined by a common factor. Therefore SIMD and GP datasets will be joined by data zone, and GP and prescriptions datasets will be joined by practice code (this is under the assumption that the practice is in the datazone). The SIMDs of the populations analysed are hence determined by the location of the GP practice that they are patients of. Residuals represent the number of prescriptions per person per GP Practice.

```{r joins, message=FALSE}
prescriptions_gp_join <- left_join(inhaler_prescriptions, gp_practices)
simd_prescriptions_gp_join <- full_join(prescriptions_gp_join, simd2020) 

# new variable is created calculating the number of combined prescriptions per person per gp practice to use are a comparative variable in analysis
prescriptions_pp_analysis <-  simd_prescriptions_gp_join %>% 
  group_by(practice_name) %>%
  summarise(prescriptions_pp_pgp = sum(number_of_paid_items)/max(practice_list_size), simd2020v2country_decile = first(simd2020v2country_decile), gp_cluster = first(gp_cluster),
    .groups = 'drop')
```

## Figure 1

```{r initial plot}
prescriptions_pp_analysis %>% 
# Some practice codes are dummy practice codes, therefore don't have an associated with SIMD. Data therefore cannot be analysed so is removed removed
# Homeless Health and Resource Services was an outlier in SIMD group 4, it is not representative of the SIMD of patients and was therefore removed
  filter(simd2020v2country_decile != "NA", practice_name != "Homeless Health & Resource Services") %>% 
# SIMD decile is used in analysis as it shows the most variation between groups (rather than quintile) so patterns will be easier to visualise
  ggplot(aes(x = simd2020v2country_decile, y = prescriptions_pp_pgp, fill = simd2020v2country_decile)) +
  geom_boxplot() +
  scale_fill_manual(values = c("1" = "#820202", "2" = "#981313", "3" = "#B71D1D", "4" = "#CC3939", "5" = "#DD6262", "6" = "#E27E7E", "7" = "#E79797", "8" = "#EBB3B3", "9" = "#EBB3B3", "10" = "#F3DBD8" )) +
  guides(fill="none") +
  labs(
    x = "SIMD Country Decile",
    y = "Number of Items Dispensed per Person per GP Practice",
    title = "Correlation of SIMD to Inhaler Prescriptions",
    subtitle = "SIMD is shown on a scale of 1 (most deprived) to 10 (least deprived)
Residuals represent the number of prescriptions per person per GP Practice",
    caption = "Source: Public Health Scotland 2020"
  ) +
  theme_minimal(base_size = 13)
```

Figure 1 shows that there is a general negative trend in SIMD to number of prescriptions per person. This data aligns with our initial hypothesis that living standards could be negatively affecting instances of chronic asthma.

# Geographical Analysis

In order to evaluate this relationship more closely, we can analyse data on a geographical level. Qualitative analysis can be done to assess the average number of prescriptions per healthboard whilst looking at the SIMD of individual GP practices within them. This is shown in Figure 2.

## Figure 2

```{r loading map data, message=FALSE}
# reading gp gractice location shapefile
gp_practice_shp <- st_read(here("data", "pub_gpprac.shp"), quiet = TRUE) %>% 
  clean_names() %>% 
  select(postcode, geometry)

#extracting postcodes as common factor between datasets
gp_postcodes <- gp_practices %>% 
  select(practice_name, postcode, gp_cluster, hb)

#giving each gp practice its postcode
gp_simd_postcodes <- full_join(prescriptions_pp_analysis, gp_postcodes)

#combibining data with shapefile
gp_simd_point_shp <- left_join(gp_practice_shp, gp_simd_postcodes, relationship = "many-to-many") %>% 
  filter(!is.na(simd2020v2country_decile))

#reading healthboard file and changing variable names for continuity
healthboard_shp <- st_read(here("data", "NHS_healthboards_2019.shp"), quiet = TRUE) %>% 
  clean_names() %>% 
  rename(hb = hb_code)

#creating a summary of prescriptions per healthboard 
healthboard_summary <- gp_simd_postcodes %>% 
    group_by(hb) %>% 
  summarise(prescriptions_pp_phb = mean(prescriptions_pp_pgp, na.rm = TRUE), simd2020v2country_decile = first(simd2020v2country_decile))

#joining summary data with shapefile
healthboard_analysis_shp <- full_join(healthboard_summary, healthboard_shp) 

```

```{r map}
#creating an overlay of GP SIMDs against healthboard average prescriptions
ggplot() +
  geom_sf(data = healthboard_analysis_shp, aes(fill = prescriptions_pp_phb, geometry = geometry), color = "black", size = 0.1) +
  geom_point(data = gp_simd_point_shp, aes(x = st_coordinates(geometry)[, 1], 
                                           y = st_coordinates(geometry)[, 2], 
                                           color = as.numeric(simd2020v2country_decile)), 
             size = 0.5) +
  scale_fill_distiller(name = "Average Prescriptions") +
  scale_color_gradient(low = "#B81010",
                       high = "#F3DBD8",
                       name = "SIMD Decile") +
    labs(
    title = "Geographic correlation of SIMD to Average Number of Prescriptions",
    subtitle = "'SIMD Decile' is shown by the location of GP Practices
'Average Prescriptions' is the mean prescriptions per person per Healthboard",
    caption = "Source: Public Health Scotland 2020",
    x = "",
    y = ""
    ) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank() 
  )
```

Health boards with higher average prescriptions (EG) appear to have a a high number of GP Practices with a high SIMD. Additionally, health board with a lower average number of prescriptions (EG) have a significant number of GP Practices with a lower SIMD score. Along with the previous findings in Figure 1, this would suggest that SIMD is affecting the number of prescriptions for preventative inhalers on a smaller geographical level. This would be logical as even neighboring data zones can have drastically different SIMD scores. Together, this indicates that individual living standards really could be having a significant impact on the prevalence chronic asthma.

# Personalised Athma Action Plan (PAAP) Resource Implementation

PAAPs are written guides that help patients with Asthma manage their condition with personal instructions in how to respond to different situations. These are very helpful for patients with chronic asthma, but they require a lot of time to put together. By assessing which GP practices have the most cases per person, resources could be better allocated to give the most affected people PAAPs, through methods such as increased GP and nurse training.

## Figure 3

```{r}
prescriptions_pp_analysis %>% 
  slice_max(prescriptions_pp_pgp, n = 10) %>% 
  select(practice_name, gp_cluster, prescriptions_pp_pgp, simd2020v2country_decile) %>%
  gt() %>% 
  tab_header(title ="GP Practices with the Highest Prescription Rates for Preventative Inhalers in Scotland", 
             subtitle = "Data from Public Health Scotland (2020)") %>% 
  cols_label(practice_name = md("**Practice Name**"),
             gp_cluster = md("**GP Cluster**"),
             prescriptions_pp_pgp = md("**Prescriptions Per Person**"),
             simd2020v2country_decile = md("**SIMD**")) %>% 
    cols_align(align = "center",
             columns = c(prescriptions_pp_pgp, simd2020v2country_decile)) %>% 
  fmt_number(columns = prescriptions_pp_pgp, decimals = 2) %>% 
  tab_style(
    style = cell_fill(color = "#FFBBA7"),
    locations = cells_body(
      rows = practice_name %in% c("Homeless Health & Resource Services",
                                  "Challenging Behaviour General Practice",
                                  "Challenging Behaviour Practice"))
  )
```

Figure 3 highlights the top 10 GP Practices in Scotland with the most prescriptions for inhalers per person, therefore these practices would likely benefit the most from increased PAAP resource allocation.
GP Cluster data shows that that 40% of these practices reside in the North East of Scotland, all with SIMD ratings of 1 or 2. This infers that this area in particular would benefit from increased aid in resource allocation, but also that it should be an area of focus for new imporved housing policy implementation.

# Conclusion
There is a clear correlation between SIMD and the number of prescriptions for drugs used in preventative asthma inhalers. Quantitative and qualitative anlaysis, in Figures 1 and 2 respectively, show that this correlation is confined in small geographical areas, suggesting that individual living standards could play a major role. Confounding factors of chronic asthma and low SIMD of homes include poor ventilation, presence of mould and poor heating, suggesting that these areas that could benefit from further policy implementation, particularly  in social housing in the North East of Scotland, as demonstrated in Figure 3.

## Limitations

Figure 2 is limited with the areas used in geographical anaylsis of prescriptions per person. As there are only 14 healthboards, datapoints within them are particularly varied, meaning that the means extracted are not particularly valuable. In the future, gographical analysis by GP Cluster would be beneficial to see finer trends as only 5-8 GP practices are within each, but shapefile data for GP clusters is currently unavailable.


Figure 3 highlights that not every GP practice is representative of their patients' SIMD status, highlighted by the rows in orange. This is logical as only areas with sufficient resources in excess can care for deprived populations, such as those that are homeless. However, this does not change the conclusion that those particular GP Practices would require more aid in giving out PAAPs to their patients. Furthermore, because of the usually defined catchment areas of most GP Practices, this flaw likely only affects a small number of practices, which does not significantly affect the rest of the analysis. This is corroborated as only one data point was removed for analysis in Figure 1 and results were still conclusive.

Perhaps the most important limitation is the assumption that SIMD is directly correlated to living standards; SIMD is a diverse measure of deprivation meaning that the two cannot be directly correlated. Although there is evidence in literature for this relationship, we cannot make the conclusion that living standards are affecting the prevalence of chronic asthma with confidence. Future analysis would benefit from being done per data zone, with each one being given a score for the average living standards for the households in the area, including mold level, ventilation and heating, along with information about which GP Practice each household attends.


# References

Douglas, A.P., Smibert, Olivia.C., Bajel, A., Halliday, C.L., Lavee, O., McMullan, B., Yong, M.K., Hal, S.J., Chen, S.C. ‐A., Slavin, M.A., Thursky, K.A., Roberts, J.A., Worth, L.J., Chang, C.C., Chen, S.C., Morrissey, C.O., Blyth, C.C. and Khanina, A. (2021). Consensus guidelines for the diagnosis and management of invasive aspergillosis, 2021. Internal Medicine Journal, 51(S7), pp.143–176. <doi:https://doi.org/10.1111/imj.15591>.

Mousavi, B. (2016). Aspergillus species in indoor environments and their possible occupational and public health hazards. Current Medical Mycology, 2(1), pp.36–42. <doi:https://doi.org/10.18869/acadpub.cmm.2.1.36>.

The Scottish Parliament (2023). Local Government, Housing and Planning Committee: Damp and Mould in Social and Private Rented Housing. [online] The Scottish Parliament. Available at: <https://www.parliament.scot/api/sitecore/CustomMedia/OfficialReport?meetingId=15285>

.Vincent, P. (2024). Asthma Inhalers. [online] Patient.info. Available at: <https://patient.info/chest-lungs/asthma-leaflet/asthma-inhalers>.

World Health Organization (2024). Asthma. [online] World Health Organisation. Available at: <https://www.who.int/news-room/fact-sheets/detail/asthma>.

### Generative AI Disclaimer

Generative AI has been used in this project for troubleshooting errors in code only
