---
title: "Assessment"
author: "Eleanor Harrison"
date: "2024-11-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Asthma is a common condition worldwide that can be caused by prolonged indoor exposure to mould (World Health Organization, 2024). Currently, there is a lack of accurate data on the prevalence of mould in social housing, potentially making this association more serious than expected (The Scottish Parliament, 2023).

This exploratory data analysis will investigate whether living standards have an impact on the prevalence of chronic asthma. To do this, living standards will be defined by SIMD and prevalence of chronic asthma will be measured by the number of prescriptions of common preventer inhalers as these are often prscribed when long or short acting bronchodilators are ineffective, suggesitng sever chronic asthma.

## Data chosen for Analysis

The prescriptions assessed are those for salbutamol, beclometasone, ciclesonide, fluticasone and mometasone as they are the most commonly prescribed preventative inhalers (Vincent, 2024)

SIMD data is released every 4 years, the dataset used will be the most recent data from 2020.

Prescriptions data used will be from 2020 to correspond with SIMD year in order to accurateky compare the two.

Data will be analysed by datazone as healthboard would be harder to show trends due to large variation in SIMD within them.

```{r libraries, include=FALSE}
library(tidyverse)
library(ggplot2)
library(janitor)
library(dplyr)
library(knitr)
library(kableExtra)
library(here)
```

### Loading Datasets

```{r loading data}
# loading datasets with desired variables selected
# some variables have been renamed for continuity between datasets
# clean names also used for continuity
simd2020 <- read_csv("data/simd2020.csv") %>% 
  clean_names() %>% 
  select(data_zone, simd2020v2country_decile) %>% 
  mutate(simd2020v2country_decile = as.factor(simd2020v2country_decile))

gp_practices <- read_csv("data/practice_contactdetails_oct2020-open-data.csv") %>% 
  rename(practice_name = GPPracticeName) %>% 
  clean_names() %>% 
  select(practice_code, data_zone)

# prescriptions data taken from jan-dec 2020, 
files <- list.files(here("data", "data_2020"), pattern = "csv")
all_prescriptions_2020 <- files %>% 
  map_dfr(~read_csv(here("data", "data_2020", .)))
```

```{r filtering}
# filtering prescriptions for desired drugs and selecting desired columns
aspergillosis_prescriptions <- all_prescriptions_2020 %>% 
  rename(HB = HBT) %>%
  rename(practice_code = GPPractice) %>% 
  clean_names() %>% 
  select(practice_code, bnf_item_description, number_of_paid_items)%>% 
  filter(str_detect(bnf_item_description, "SALBUTAMOL|BECLOMETASONE|CICLESONIDE|FLUTICASONE|MOMETASONE"))
```

The three datasets used for this analysis differ by the areas given therefore cannot be joined by a common factor. Therefore SIMD and GP datasets will be joined by data zone, and GP and prescriptions dataset will be joined by practice code (this is under the assumptiosn that the practice is in the datazone).

```{r joins}
prescriptions_gp_join <- left_join(aspergillosis_prescriptions, gp_practices)
simd_prescriptions_gp_join <- full_join(prescriptions_gp_join, simd2020)
# some practices dont have a corresponding datazones - find later
```


```{r initial plot}
simd_prescriptions_gp_join %>% 
  filter() %>% 
  group_by(simd2020v2country_decile) %>% 
  ggplot(aes(x = simd2020v2country_decile, y = number_of_paid_items)) +
  geom_col() +
  labs(
    x = "SIMD Country Decile",
    y = "Number of Items Dispensed"
  )+
  theme_minimal()
```

 


